---
title: "Your title here"
author:
  - name: 
    affiliation: 
date: "`r format(Sys.time(), '%B %d, %Y')`" 
output:
  html_document:
    code_folding: show
    highlights: bg-maroon
    number_sections: yes
    theme: journal
    toc: yes
    toc_float: yes
---
# Information
> Script to run UMAP, clustering, and automatic graph generation from flow cytometry data 
> Written by Rebecca Voogt in March 2020 in R version 3.5.3, updated in R 4.0.2   
> Acknowledgement and credit for code and ideas from Lei Huang (UChicago Bioinformatics Core), Young Yoon (Sperling Lab), UIC edgeR workshop, as well as various Internet sources (links at the end). Funding provided by Anne Sperling  
> This version was finished March 29, 2021

# Instructions
## Instructions Part 1: Intro
> FIRST, save this document in the folder with your data with a unique name. DO NOT overwrite the template.

The first time through this document, please follow the instructions starting with the next line. After you have run this chunk by chunk and don't get any errors, please knit it (click the knit button at the top with the yarn ball icon) so that your work is saved.

> NOTE: All figures will be auto-saved to the folder you choose, EXCEPT the pseudocolor UMAP. You will need to click the camera icon at the left hand side of the strip of navigation tools after the graph has been produced and manually save it.

> NOTE: If this is your first time running this script, please make sure that you have installed all of the packages in the chunk below.

Click the green triangle in the top right of the first grey box only if this is your first time running the script on this computer. It will take some time to install all the packages.

Click the green triangle in the top right of grey box. Follow the next instructions.

```{r Setup2, message=FALSE, warning=FALSE, results='hide'}
# clean up environment
rm(list=ls())

# import needed packages
suppressPackageStartupMessages(library(data.table))
suppressPackageStartupMessages(library(plyr))
suppressPackageStartupMessages(library(dplyr))
suppressPackageStartupMessages(library(tidyr))
suppressPackageStartupMessages(library(ggplot2))
suppressPackageStartupMessages(library(umap))
suppressPackageStartupMessages(library(circlize))
suppressPackageStartupMessages(library(ComplexHeatmap))
suppressPackageStartupMessages(library(FlowSOM))
suppressPackageStartupMessages(library(flowCore))
suppressPackageStartupMessages(library(dmm))
suppressPackageStartupMessages(library(factoextra))
suppressPackageStartupMessages(library(ggpubr))
suppressPackageStartupMessages(library(here))
suppressPackageStartupMessages(library(lubridate))
suppressPackageStartupMessages(library(tidyverse))
suppressPackageStartupMessages(library(plotly))
suppressPackageStartupMessages(library(ggrepel))
suppressPackageStartupMessages(library(limma))
suppressPackageStartupMessages(library(CATALYST))
suppressPackageStartupMessages(library(RColorBrewer))

# set seed for reproducibility
set.seed(2936)
```
## Instructions
### Choose options
Fill in the options to the right of the arrow inside the quotes if applicable in the gray box below with 

experimental condition (p.condition) must be exactly as on the metadata sheet column name
name of group 1 (p.group1) must be exactly as on the metadata sheet
name of group 2 (p.group2) must be exactly as on the metadata sheet
name of group 3 (p.group3) must be exactly as on the metadata sheet
number of cells per sample (p.cells)

number of clusters you want to test in the flowSOM (p.testclusternumber)
OR number of clusters you want to have in the flowSOM (p.absclusternumber)

the full path to where your files are located (p.workingdirectory)
the name of your file (p.file)
the name of your metadata CSV file (p.metadata)
unique experiment name (p.exp_name)
colors for the cluster graph, make sure there is at least as many as max clusters to test or absolute number of clusters specified (p.colors)

Click the green triangle in the top right of the next two gray boxes. Follow the next instructions.

```{r, VarSet1, echo=FALSE, warning=FALSE}
p.condition <- "group"
p.group1 <- "baseline"
p.group2 <- "group 2"
p.group3 <- "group 3"
p.cells <- 300
p.levels <- c("baseline", "group 2", "group 3", ordered = TRUE)
p.testclusternumber <- 30
p.absclusternumber <- 15
p.workingdirectory <- "path/to/file"
p.meta <- "./metadata.csv"
p.exp_name <- "unique expt name"
```

## Setup continued
```{r, Setup3, echo=FALSE, warning=FALSE}
# set working directory
d.all <- list.files(path = p.workingdirectory,
         pattern="export_.", 
         full.names = T) %>% 
   map_df(function(x) read.csv(x) %>% mutate(file_name=gsub(".csv","",basename(x)))) 

setwd(p.workingdirectory)

d.metadata <- read_csv(p.meta)

d.all <- merge(d.all, d.metadata,
        by = c("file_name"))

colnames(d.all)

```
## Instructions Part 3: Setting variables
To see a list of all possible column names either 1) look above  or 2) click on d.all in the list of variables in the environment in the top right corner of RStudio. It will open in a tab in the top left of RStudio. Look closely at the column names, and use them to fill in the variables to the right of the arrow inside the c( ) in the gray box below. Be sure to type them exactly as in the column headers, underscores, periods, and dashes included. Each column name needs to be in "quotes" and separated by commas. When you're done, click in the next gray box, and in the top right of this pane click "run" > "run all chunks below".

variables you want to be included in the analyses. MAKE SURE "sample_id" and all of your experimental columns are in the list (p.vars)

p.mds should be p.vars minus your grouping variable

p.umap should be p.mds minus your sample_id variable

```{r, VarSet2, echo=FALSE, warning=FALSE}
p.vars <- c("sample_id", "group", "GP38", "CD31", "MHCII", "EPCAM")

p.mds <- c("sample_id", "GP38", "CD31", "MHCII", "EPCAM")

p.umap <- c("GP38", "CD31", "MHCII", "EPCAM")

```

Please do not type below this line!

```{r, printVars}
print("p.vars variable includes:")
print(p.vars)
print("p.mds variable includes:")
print(p.mds)
print("p.umap variable includes:")
print(p.umap)
print("number of cells per sample:")
print(length(rownames(d.all))/length(unique(d.all$sample_id)))
```

# MDS
```{r warning=FALSE, MDS, echo=FALSE}
d.metadata$group <- factor(d.metadata$group)

# make a duplicate dataframe for later
d.pca <- d.all

# set working directory, and assign group/treatment for each cell per sample
setwd(p.workingdirectory)


# Subset only the marker intensity and sample id columns
eset <- d.pca[, colnames(d.pca) %in% p.mds]


# take median intensity of the cells belonging to individual sample
expr_median_sample_tbl <- eset %>%
  group_by(sample_id) %>%
  dplyr::summarize_all(funs(median))

# convert to matrix
expr_median_sample <- t(expr_median_sample_tbl[, -1])
colnames(expr_median_sample) <- expr_median_sample_tbl$sample_id

treatmds <- d.metadata[,p.condition]
colnames(treatmds) <- "condition"

# if samples have other meta data, e.g. 7 samples belong to two different groups
# you can define the color of the groups.
color_conditions <- brewer.pal(n = length(levels(d.metadata$group)), name = "Set1")[1:length(levels(d.metadata$group))]
names(color_conditions) <- levels(d.metadata$group)

# run plotMDS() by setting plot = FALSE (the function is from limma package)
mds <- limma::plotMDS(expr_median_sample, plot = FALSE)

# create a data frame to be used as input for ggplot
ggdf <- data.frame(MDS1 = mds$x,
                   MDS2 = mds$y,
                   sample_id = colnames(expr_median_sample),
                   disease = treatmds)

# make and save ggplot of MDS colored by sample
p <- ggplot(ggdf, aes(x = MDS1, y = MDS2, color = sample_id)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_label_repel(aes(label = sample_id), show.legend = FALSE) +
  theme_bw() +
  coord_fixed(ratio = 3)

plot(p)

ggsave(p, filename=paste("MDS_",p.exp_name,".png", sep = ""), bg = "transparent")

# make and save ggplot of MDS colored by group
e <- ggplot(ggdf, aes(x = MDS1, y = MDS2, color = condition)) +
  geom_point(size = 3, alpha = 0.8) +
  geom_label_repel(aes(label = sample_id), show.legend = FALSE) +
  theme_bw() +
  scale_color_manual(values = color_conditions) +
  coord_fixed(ratio = 3)

plot(e)

ggsave(e, filename=paste("MDS_by_disease_",p.exp_name,".png", sep = ""), bg = "transparent")

```

# UMAP
```{r, UMAP, echo=FALSE, warning=FALSE}
setwd(p.workingdirectory)
# make a dataframe with only relevant columns
d.data <- d.all[,p.umap]

d.labels <- d.all$sample_id
# umap
data.umap <- umap(d.data)

# start a new dataframe with UMAP generated coordinates
d.umap <- data.frame(data.umap$layout)
d.umap <- cbind(d.umap, d.labels)
colnames(d.umap)[1] <- "UMAP1"
colnames(d.umap)[2] <- "UMAP2"

# plot overall UMAP
w <- ggplot(d.umap,aes(y=UMAP2, x=UMAP1)) + 
  geom_point(size= 0.005, color="black", shape = 20)+
  labs(title="UMAP")+
  coord_fixed(ratio = 1)
plot(w)
ggsave(w, filename=paste("UMAP_",p.exp_name,".png", sep = ""), bg = "transparent")

# make and save a ggplot of the umap for each group
d.manyumap <- d.umap
colnames(d.manyumap)[3] <- "sample_id"
d.manyumap <- merge(d.manyumap, d.metadata,
        by = "sample_id")

d.manyumap[4] <- NULL

d.group1 <- filter(d.manyumap,
                 group == p.group1)

d.group2 <- filter(d.manyumap,
                 group == p.group2)

d.group3 <- filter(d.manyumap,
                 group == p.group3)

h <- ggplot(d.manyumap,aes(y=UMAP2, x=UMAP1)) + 
  geom_point(size= 0.05, color="black", shape = 20)+ 
  facet_wrap(~group)+
  labs(title="UMAP Plotted by Group")
plot(h)
ggsave(h, filename=paste("UMAP_plotted_by_group_",p.exp_name,".png", sep = ""), bg = "transparent")

i <- ggplot(d.group1,aes(y=UMAP2, x=UMAP1)) + 
  geom_point(size= 0.05, color="black", shape = 20)+ 
  labs(title=paste("UMAP ",p.group1))+
  coord_fixed(ratio = 1)
plot(i)
ggsave(i, filename=paste("UMAP_",p.group1,"_",p.exp_name,".png", sep = ""), bg = "transparent")

j <- ggplot(d.group2,aes(y=UMAP2, x=UMAP1)) + 
  geom_point(size= 0.05, color="black", shape = 20)+ 
  labs(title=paste("UMAP ",p.group2))+
  coord_fixed(ratio = 1)
plot(j)
ggsave(j, filename=paste("UMAP_",p.group2,"_",p.exp_name,".png", sep = ""), bg = "transparent")

k <- ggplot(d.group3,aes(y=UMAP2, x=UMAP1)) + 
  geom_point(size= 0.05, color="black", shape = 20)+ 
  labs(title=paste("UMAP ",p.group3))+
  coord_fixed(ratio = 1)
plot(k)
ggsave(k, filename=paste("UMAP_",p.group3,"_",p.exp_name,".png", sep = ""), bg = "transparent")

# make and save a ggplot of the umap for each sample
g <- ggplot(d.umap,aes(y=UMAP2, x=UMAP1)) + 
  geom_point(size= 0.05, color="black", shape = 20)+ 
  facet_wrap(~d.labels)+
  labs(title="UMAP Plotted by Sample")
plot(g)
ggsave(g, filename=paste("UMAP_plotted_by_sample_",p.exp_name,".png", sep = ""), bg = "transparent")

```

# FlowSOM Clustering
```{r FlowSOM, echo=FALSE, warning=FALSE}
setwd(p.workingdirectory)
# make a flowframe object from the imported data
d.flowframe <- flowFrame(as.matrix(d.data))

# run the flowSOM
data.flowsom <- FlowSOM(d.flowframe, compensate = FALSE, transform = FALSE, scale = TRUE, silent = FALSE, colsToUse = p.umap, maxMeta = p.testclusternumber, seed = 2936)
# plot the tree map and save it
pdf(paste("FlowSOM_tree_",p.exp_name,".pdf"), width = 8, height = 8)
PlotStars(data.flowsom[[1]])
dev.off()
PlotStars(data.flowsom[[1]])
# make a new object with the cluster identification
flowSOM_cluster <- data.flowsom[[2]][data.flowsom[[1]]$map$mapping[,1]]
# copy the dataset with only variables of interest and then add cluster ID as a new column
mydata <- d.data
mydata <- data.frame(mydata, flowSOM_cluster)
d.umap <- data.frame(d.umap, flowSOM_cluster)

# Determine number of points in each cluster per sample
d.points <- dplyr::select(d.umap, 
                          c(d.labels,flowSOM_cluster))
colnames(d.points)[1] <- "sample_id"
colnames(d.points)[2] <- "cluster"
max.sample <- max(d.points$sample_id)
max.cluster <- max(as.numeric(d.points$cluster))

catvars <- c("sample_id","cluster")
d.points[catvars] <- lapply(d.points[catvars], factor)

count.table <- d.points %>%
  group_by(sample_id, cluster) %>%
  dplyr::summarize(stat = n()) %>%
  spread(cluster, stat, fill = 0)

samplenums <- count.table[1]

count.table <- count.table[-c(1)]
```

# Plot the UMAP by sample overlaid with flowSOM cluster and save it
```{r, UMAPcluster, echo=FALSE, warning=FALSE}
setwd(p.workingdirectory)
# plot and save overall umap colored by flowSOM cluster
b <- ggplot(d.umap,aes(y=UMAP2, x=UMAP1)) + 
  geom_point(aes(color=factor(flowSOM_cluster)),size=1, shape = 20)+
  labs(title="UMAP")+
  scale_color_manual(values = c("#3281a8", "#999999", "#000000", "#09851e", "#9b30c2", 
                                "#31e0cc", "#d4a224", "#ed1202", "#47e31b", "#a23929", 
                                "#abcdef", "#ffaf2e", "#909000", "#f7f42f", "#cdcdcd"))
plot(b)
ggsave(b, filename=paste("UMAP_overlaid_with_FlowSom_Cluster",p.exp_name,".png", sep = ""), bg = "transparent")

d.manyflowsom <- d.umap
colnames(d.manyflowsom)[3] <- "sample_id"
d.manyflowsom <- merge(d.manyflowsom, d.metadata,
        by = "sample_id")

d.manyflowsom[5] <- NULL

# plot and save umap colored by flowSOM cluster and split into groups
m <- ggplot(d.manyflowsom,aes(y=UMAP2, x=UMAP1)) + 
  geom_point(aes(color=factor(flowSOM_cluster)),size=1, shape = 20)+ 
  facet_wrap(~group)+
  labs(title="UMAP by Group and Colored by FlowSOM Cluster")+
  scale_color_manual(values = c("#3281a8", "#999999", "#000000", "#09851e", "#9b30c2", 
                                "#31e0cc", "#d4a224", "#ed1202", "#47e31b", "#a23929", 
                                "#abcdef", "#ffaf2e", "#909000", "#f7f42f", "#cdcdcd"))
plot(m)
ggsave(m, filename=paste("UMAP_plotted_by_group_and_cluster_",p.exp_name,".png", sep = ""), bg = "transparent")

d.group11 <- filter(d.manyflowsom,
                  group == p.group1)
d.group21 <- filter(d.manyflowsom,
                  group == p.group2)
d.group31 <- filter(d.manyflowsom,
                  group == p.group3)

i <- ggplot(d.group11,aes(y=UMAP2, x=UMAP1)) + 
  geom_point(aes(color=factor(flowSOM_cluster)),size=1, shape = 20)+ 
  labs(title=paste("UMAP ",p.group1," Colored by FlowSOM Cluster"))+
  scale_color_manual(values = c("#3281a8", "#999999", "#000000", "#09851e", "#9b30c2", 
                                "#31e0cc", "#d4a224", "#ed1202", "#47e31b", "#a23929", 
                                "#abcdef", "#ffaf2e", "#909000", "#f7f42f", "#cdcdcd"))
plot(i)
ggsave(i, filename=paste("UMAP_",p.group1,"_plotted_by_group_and_cluster_",p.exp_name,".png", sep = ""), bg = "transparent")

j <- ggplot(d.group21,aes(y=UMAP2, x=UMAP1)) + 
  geom_point(aes(color=factor(flowSOM_cluster)),size=1, shape = 20)+
  labs(title=paste("UMAP ",p.group2," Colored by FlowSOM Cluster"))+
  scale_color_manual(values = c("#3281a8", "#999999", "#000000", "#09851e", "#9b30c2", 
                                "#31e0cc", "#d4a224", "#ed1202", "#47e31b", "#a23929", 
                                "#abcdef", "#ffaf2e", "#909000", "#f7f42f", "#cdcdcd"))
plot(j)
ggsave(j, filename=paste("UMAP_",p.group2,"_plotted_by_group_and_cluster_",p.exp_name,".png", sep = ""), bg = "transparent")

r <- ggplot(d.group31,aes(y=UMAP2, x=UMAP1)) + 
  geom_point(aes(color=factor(flowSOM_cluster)),size=1, shape = 20)+
  labs(title=paste("UMAP ",p.group3," Colored by FlowSOM Cluster"))+
  scale_color_manual(values = c("#3281a8", "#999999", "#000000", "#09851e", "#9b30c2", 
                                "#31e0cc", "#d4a224", "#ed1202", "#47e31b", "#a23929", 
                                "#abcdef", "#ffaf2e", "#909000", "#f7f42f", "#cdcdcd"))
plot(r)
ggsave(r, filename=paste("UMAP_",p.group3,"_plotted_by_group_and_cluster_",p.exp_name,".png", sep = ""), bg = "transparent")

# plot and save umap colored by flowSOM cluster and split into each sample
k <- ggplot(d.umap,aes(y=UMAP2, x=UMAP1)) + 
  geom_point(aes(color=factor(flowSOM_cluster)),size=1, shape = 20)+ 
  facet_wrap(~d.labels)+
  labs(title="UMAP by Sample and Colored by FlowSOM Cluster")+
  scale_color_manual(values = c("#3281a8", "#999999", "#000000", "#09851e", "#9b30c2", 
                                "#31e0cc", "#d4a224", "#ed1202", "#47e31b", "#a23929", 
                                "#abcdef", "#ffaf2e", "#909000", "#f7f42f", "#cdcdcd"))
plot(k)
ggsave(k, filename=paste("UMAP_plotted_by_sample_and_cluster_",p.exp_name,".png", sep = ""), bg = "transparent")

```

# UMAP Pseudocolor Graph
```{r, UMAP.Pseudocolor, echo=FALSE, warning=FALSE}
setwd(p.workingdirectory)
UMAP1 <- d.umap$UMAP1
UMAP2 <- d.umap$UMAP2
# use plotly package to plot a density/pseudocolor map of umap. lighter color indicates more cells there
s <- subplot(
  plot_ly(x = UMAP1, type = "histogram"),
  plotly_empty(),
  plot_ly(x = UMAP1, y = UMAP2, type = "histogram2dcontour"),
  plot_ly(y = UMAP2, type = "histogram"),
  nrows = 2, heights = c(0.2, 0.8), widths = c(0.8, 0.2), margin = 0,
  shareX = TRUE, shareY = TRUE, titleX = TRUE, titleY = TRUE
)
x <- list(title="UMAP1", center = TRUE)
y <- list(title="UMAP2", center = TRUE)

s <- plot_ly(x = UMAP1, y = UMAP2, type = "histogram2dcontour") %>%
  layout(xaxis = x, yaxis = y)
s
```

# UMAP with marker intensity color overlay
```{r, colorbyfluor, echo=FALSE, warning=FALSE}
d.fluormap <- d.data
d.fluormap$UMAP1 <- d.umap$UMAP1
d.fluormap$UMAP2 <- d.umap$UMAP2

# for each color used in umap and flowSOM analysis, plot and save a graph of where that marker is "expressed"
for(i in 1:length(p.umap)){
o <- ggplot(d.fluormap, aes(x=UMAP1, y= UMAP2, color=d.fluormap[,i])) + 
  geom_point(size=1, shape = 20)+ 
  labs(title=paste("UMAP Marker Color Overlay - ",colnames(d.fluormap)[i]))+
  scale_color_gradient2(midpoint = 500, low="blue",mid="grey", high="red")

plot(o)
ggsave(o, filename=paste("UMAP_plotted_by_marker_",colnames(d.fluormap)[i],"_",p.exp_name,".png", sep = ""), bg = "transparent")
}

```


# Heatmap
```{r, Heatmap, echo=FALSE, warning=FALSE}
# this block makes a heatmap with z-scores indicating where cells from each cluster are coming from
# credit: UIC edgeR workshop for color gradient

setwd(p.workingdirectory)
# make a table of counts per sample per cluster
count.table.t <- data.frame(t(count.table))
colnames(count.table.t) <- t(samplenums)
# normalize the table
count.norm.z <- t(scale(t(count.table.t)))

# sum the number of cells in each cluster
clustersize <- colSums(count.table)
data.frame(clustersize)
# plot a barplot of cluster size on the heatmap axis
rowha <- rowAnnotation(Cells_in_Cluster = anno_barplot(clustersize))

# plot the heatmap. once to save and once to display in knitted html document
col_fun <-colorRamp2(c(-3,0,3),c("blue","white","red"))
pdf(paste("Heatmap_",p.exp_name,".pdf", sep = ""), width = 8, height = 8)
Heatmap(count.norm.z,show_row_names = T, name="z-score", col= col_fun, row_title = "cluster", left_annotation = rowha, column_title = "sample", use_raster = FALSE)
dev.off()

Heatmap(count.norm.z,show_row_names = T, name="z-score", col= col_fun, row_title = "cluster", left_annotation = rowha, column_title = "sample", use_raster = FALSE)
rm(count.table.t, count.norm.z)
```

# Cluster phenotypes (avg expression per marker)
```{r, ClusterPheno, echo=FALSE, warning=FALSE}
# this block makes a graph for each cluster showing each marker's expression
setwd(p.workingdirectory)
# make a dataframe by adding cluster number as a feature for each cell
cluster.data <- cbind(d.data, mydata$flowSOM_cluster)
colnames(cluster.data)[(strtoi(length(p.umap)))+1] <- "cluster"

# calculate average expression of each marker for each cluster
cluster.avg <- aggregate(cluster.data[,1:strtoi(length(p.umap))], list(cluster.data$cluster), mean)
cluster.avg <- cluster.avg[-c(1)]
markerlist <- colnames(cluster.avg)
cluster.avg <- t(cluster.avg)
cluster.avg <- data.frame(cluster.avg)
colnames(cluster.avg) <- seq(1,strtoi(max.cluster))

clusternum <- max.cluster
d.graphdata <- cluster.avg

setDT(d.graphdata, keep.rownames = TRUE)
colnames(d.graphdata)[1] <- "Parameter"
colnames(d.graphdata)[2:(clusternum+1)] <- paste("Cluster", seq(1:clusternum), sep = "")

# code for line below from Young Me Yoon
# determine max value for each marker
d_max <- d.graphdata %>% summarise_if(is.numeric, max, na.rm=TRUE)
# determine max value overall
v_max <- max(t(d_max))

# use tidyr::gather to make long form of cluster parameter info for graphing
d.graph.long <- gather(d.graphdata, cluster, value, -Parameter, factor_key = TRUE)

# graph all parameter information for all clusters and save it
q <- ggplot(d.graph.long, aes(fill=cluster, y=value, x=Parameter))+
  geom_bar(position="dodge", stat="identity")+
  theme(axis.text.x = element_text(angle=45, hjust=1))+
  theme(legend.position = "bottom")

plot(q)

ggsave(q, filename=paste("All_cluster_phenotypes",p.exp_name,".png", sep = ""), bg= "transparent")

# plot and save a graph for each cluster showing marker expression
d.graphdata <- data.frame(d.graphdata)
for(i in 2:(clusternum+1)){
z <- ggplot(d.graphdata, aes(x=Parameter, y= d.graphdata[,i])) + 
  geom_bar(position="dodge", stat="identity", width = 0.6)+
  coord_cartesian(ylim = c(0, v_max)) +
  labs(y="MFI", title=colnames(d.graphdata)[i])+
  theme(axis.text.x = element_text(angle=45, hjust=1))

plot(z)
ggsave(z, filename=paste(colnames(d.graphdata)[i],"_phenotype_",p.exp_name,".png", sep = ""), bg = "transparent")
}
```

# Cluster Phenotype Heatmap
```{r warning=FALSE, Clusterheatmap, echo=FALSE}
# this block makes a heatmap showing relative high or low expression for each marker for each cluster
setwd(p.workingdirectory)
p.clusterpheno <- d.graphdata
rownames(p.clusterpheno) <- p.clusterpheno[,1]
p.clusterpheno[,1] <- NULL
colnames(p.clusterpheno) <- seq(1, clusternum)
clusterpheno.z <- t(scale(t(p.clusterpheno)))

# sum the number of cells in each cluster
clustersize <- colSums(count.table)
data.frame(clustersize)
# plot a bargraph of cluster size on the heatmap axis
columnha <- HeatmapAnnotation(Cells_in_Cluster = anno_barplot(clustersize))

# plot the heatmap twice, once to save and once to display in knitted html file
col_fun <-colorRamp2(c(-3,0,3),c("blue","white","red"))
pdf(paste("Cluster_Phenotype_Heatmap_",p.exp_name,".pdf"), width = 8, height = 8)
Heatmap(clusterpheno.z, 
        column_labels = colnames(clusterpheno.z), 
        row_labels = rownames(clusterpheno.z), 
        name="z-score", 
        col= col_fun, 
        row_title = "Parameter", 
        column_title = "Cluster", 
        top_annotation = columnha,
        cluster_rows = FALSE,
        use_raster = FALSE)
dev.off()

Heatmap(clusterpheno.z, 
        column_labels = colnames(clusterpheno.z), 
        row_labels = rownames(clusterpheno.z), 
        name="z-score", 
        col= col_fun, 
        row_title = "Parameter", 
        column_title = "Cluster", 
        top_annotation = columnha,
        cluster_rows = FALSE,
        use_raster = FALSE)
```


# Graphs comparing number of cells per cluster, graphed by condition/group
```{r Conditiongraph}
setwd(p.workingdirectory)
# define experimental conditions
condition <- d.metadata$group

# make a new table to add condition
count.table.condition <- count.table
count.table.condition$testvar <- condition
count.table.condition <- as.data.frame(count.table.condition)

# reorder columns
count.table.condition <- count.table.condition[c(strtoi(max.cluster)+1, 1:strtoi(max.cluster))]

# make an average table for genotype and treatment (Young Yoon)
d.avg.condition <- count.table.condition %>% 
  group_by(testvar) %>% 
  summarise_if(is.numeric, mean, na.rm=TRUE)
d.avg.condition <- as.data.frame(d.avg.condition)

#To fix the space at the bottom, separately assign limits to bottom and top. (code from Young Yoon)
##First create a dataframe that contains max value, max*1.1, max*1.2 to use for ylimits
d_max <- count.table %>% summarise_if(is.numeric, max, na.rm=TRUE)
###Add a column in front
d_max <- data.frame(a=0, d_max)

d_max_1.2 <- d_max*1.2
d_max_1.1 <- d_max*1.1
d_max_1.3 <- d_max*1.3
d_max_1.4 <- d_max*1.4
d_max_1.5 <- d_max*1.5

# make bar graphs comparing treatment for each population using a for loop (Young Yoon)
for (i in 2:(strtoi(max.cluster)+1)){
  chartTitle1 <- paste("FlowSOM Cluster ", colnames(count.table.condition)[i])
  axislabel <- paste("# cells in FlowSOM cluster ", colnames(count.table.condition)[i])
  plt <- ggplot(data=count.table.condition,aes(x=testvar, y=count.table.condition[,i])) +
    geom_point(aes(color=testvar), size=3, 
               position=position_jitter(w=0.08),
               show.legend=FALSE) +
    geom_bar(data=d.avg.condition,
             aes(x=testvar, y=d.avg.condition[,i]), 
             stat="identity", alpha=0, color="black", width=0.75) +
    labs(y=axislabel, x= p.condition, title = chartTitle1) +
    theme_classic() +
    theme(text=element_text(size=12, color="black"),
          plot.title=element_text(size=15)) +
    theme(aspect.ratio=0.5) +
    stat_compare_means(comparisons = list(c(p.group1,p.group2), c(p.group2,p.group3), c(p.group1,p.group3)), 
                       label.y = c(d_max_1.1[,i], d_max_1.1[,i], d_max_1.2[,i]),
                       size=7) +
    stat_compare_means(method="anova",
                       label.y = d_max_1.4[,i], 
                       size=7)+
    scale_y_continuous(expand=c(0,0)) +
    expand_limits(y=d_max_1.5[,i])
  print(plt)
  ggsave(plt, filename=paste("plt_FlowSOM_Cluster_",colnames(count.table.condition)[i],p.exp_name,".png"), bg = "transparent")
}
```

# Make full dataframes and export them
```{r, Exportdata, echo=FALSE, warning=FALSE}
setwd(p.workingdirectory)
d.all <- data.frame(d.all,flowSOM_cluster)
d.all$UMAP1 <- d.umap$UMAP1
d.all$UMAP2 <- d.umap$UMAP2
d.export <- mydata
d.export$sampleid <- d.labels
d.export$UMAP1 <- d.umap$UMAP1
d.export$UMAP2 <- d.umap$UMAP2
colnames(d.umap)[3] <- "sampleid"
d.clusterexp <- count.table.condition
d.clusterexp$sampleid <- unlist(samplenums)

d.avg.cluster <- d.clusterexp[,2:max.cluster]/p.cells*100
d.avg.cluster$sampleid <- d.clusterexp$sampleid
d.avg.cluster$testvar <- d.clusterexp$testvar

write.csv(d.clusterexp, file=paste("Number_cells_per_cluster_by_sample_",p.exp_name,".csv", sep = ""),row.names = FALSE)

write.csv(d.all, file=paste("All_data_with_cluster_sampleid_UMAP_",p.exp_name,".csv", sep = ""),row.names = FALSE)

write.csv(d.export, file=paste("Parameters_used_to_cluster_with_cluster_sampleid_UMAP_",p.exp_name,".csv", sep = ""), row.names = FALSE)

write.csv(d.umap, file=paste("UMAP_coordinates_sampleid_cluster_data_",p.exp_name,".csv", sep = ""), row.names = FALSE)

write.csv(cluster.avg, file=paste("Cluster_phenotype_info_",p.exp_name,".csv", sep = ""), row.names = TRUE)

write.csv(d.avg.cluster, file=paste("Cluster_percentages_",p.exp_name,".csv", sep = ""), row.names = FALSE)
```
# Graphs and files produced
You will have  CSV files produced, the file names start with    
Number_cells_per_cluster_by_sample_ (columns are cluster numbers)  
All_data_with_cluster_sampleid_UMAP_    
Parameters_used_to_cluster_with_cluster_sampleid_UMAP_    
UMAP_coordinates_sampleid_cluster_data_    
Cluster_phenotype_info_  

You will have a pdf of the heatmap which starts with Heatmap_    
You will have a pdf of the cluster phenotype heatmap which starts with Cluster_Phenotype_Heatmap_  

You will have a pdf of the FlowSom tree which starts with FlowSOM_tree_  

You will have two pdfs of the MDS, they start with  
MDS_plot_  
MDS_plot_by...  

You will have 3 UMAP .png files  
UMAP_  
UMAP_plotted_by_sample_  
UMAP_plotted_by_sample_and_cluster_  

You will have a UMAP color overlay for each of your markers  

You will have a .png file of a bar graph of each cluster comparing the groups specified. The number of graphs will vary depending on how many clusters were made.  

You will have a .png file of a bar graph showing the mean fluorescence intensity of each color for each cluster, all should have the same scale.

# Sources 
https://community.rstudio.com/t/error-aesthetics-must-be-either-length-1-or-the-same-as-the-data-2-fill/15579/3
https://stackoverflow.com/questions/57063269/plotting-data-in-r-error-default-method-not-implemented-for-type-list
https://www.statmethods.net/input/datatypes.html
https://stackoverflow.com/questions/5620885/how-does-one-reorder-columns-in-a-data-frame
https://stackoverflow.com/questions/21982987/mean-per-group-in-a-data-frame
https://stackoverflow.com/questions/54050320/error-data-must-be-a-data-frame-or-other-object-coercible-by-fortify-no
https://www.r-graph-gallery.com/48-grouped-barplot-with-ggplot2.html
https://stackoverflow.com/questions/22359127/how-to-repeat-a-string-n-times-in-r
https://stackoverflow.com/questions/1663370/loop-over-string-variables-in-r
https://stackoverflow.com/questions/24428051/removing-display-of-row-names-from-data-frame
https://stackoverflow.com/questions/12019461/rbind-error-names-do-not-match-previous-names/12019514
https://www.edureka.co/community/53878/change-ggplot2-default-color-palette
https://jokergoo.github.io/ComplexHeatmap-reference/book/a-single-heatmap.html#heatmap-as-raster-image
https://stackoverflow.com/questions/11454333/r-numbers-from-1-to-100/11454412
https://stackoverflow.com/questions/6081439/changing-column-names-of-a-data-frame
https://www.datamentor.io/r-programming/if-else-statement/
https://www.quora.com/How-do-I-get-a-frequency-count-based-on-two-columns-variables-in-an-R-dataframe
https://www.edureka.co/community/4963/creating-an-empty-data-frame-with-only-column-names-r
https://stackoverflow.com/questions/17728540/compress-rows-with-na-in-different-columns-to-create-a-single-row
https://stat.ethz.ch/R-manual/R-devel/library/base/html/strtoi.html
https://www.guru99.com/r-data-frames.html
https://rdrr.io/cran/dplyr/man/group_by.html
https://stackoverflow.com/questions/55248086/r-dplyrs-group-by-consider-empty-groups-as-well
https://www.rdocumentation.org/packages/dmm/versions/2.1-5/topics/unfactor
https://stackoverflow.com/questions/46325145/error-with-select-function-from-dplyr
https://community.rstudio.com/t/error-cannot-allocate-vector-of-size-76-4-gb/10615/2
https://www.statmethods.net/advstats/cluster.html
https://ggplot2.tidyverse.org/reference/
https://ggplot2.tidyverse.org/reference/geom_dotplot.html
https://ggplot2.tidyverse.org/reference/geom_point.html
https://stackoverflow.com/questions/26643852/ggplot-plots-in-scripts-do-not-display-in-rstudio
https://stackoverflow.com/questions/7714677/scatterplot-with-too-many-points
https://www.r-bloggers.com/7-visualizations-you-should-learn-in-r/
https://stackoverflow.com/questions/26247429/data-manipulation-in-r-x-must-be-atomic
https://www.statmethods.net/advstats/cluster.html
https://stackoverflow.com/questions/12868581/list-all-column-except-for-one-in-r
https://stackoverflow.com/questions/4605206/drop-data-frame-columns-by-name
https://ggplot2.tidyverse.org/
https://cran.r-project.org/web/packages/umap/vignettes/umap.html
https://stackoverflow.com/questions/7070173/convert-data-frame-column-to-a-vector
https://www.researchgate.net/post/Can_anybody_explain_why_R_cannot_read_a_certain_file
https://yihui.org/knitr/options/#other-chunk-options
https://yihui.org/knitr/options/
https://rstudio.com/wp-content/uploads/2015/02/rmarkdown-cheatsheet.pdf
https://rpkgs.datanovia.com/ggpubr/reference/stat_compare_means.html
https://stackoverflow.com/questions/1330989/rotating-and-spacing-axis-labels-in-ggplot2
https://uc-r.github.io/tidyr
https://discuss.analyticsvidhya.com/t/how-to-add-a-column-to-a-data-frame-in-r/3278
https://stackoverflow.com/questions/5555408/convert-the-values-in-a-column-into-row-names-in-an-existing-data-frame-in-r
https://www.r-graph-gallery.com/142-basic-radar-chart.html
https://ggplot2.tidyverse.org/reference/lims.html
https://stackoverflow.com/questions/56866178/dplyrn-returns-error-error-n-should-only-be-called-in-a-data-context
https://stackoverflow.com/questions/29511215/convert-row-names-into-first-column
https://plotly.com/r/contour-plots/#2d-density-contour-plot
https://plotly.com/r/figure-labels/
https://plotly.com/r/plotly-fundamentals/
https://plotly.com/r/colorscales/
https://plotly.com/r/line-and-scatter/#custom-color-scales
https://plotly.com/r/reference/#contour
http://www.sthda.com/english/wiki/ggplot2-colors-how-to-change-colors-automatically-and-manually
https://ggplot2.tidyverse.org/reference/scale_colour_continuous.html
https://rstudio.com/wp-content/uploads/2015/03/ggplot2-cheatsheet.pdf
https://stackoverflow.com/questions/20415963/how-to-increase-the-size-of-points-in-legend-of-ggplot2
https://stackoverflow.com/questions/14942681/change-size-of-axes-title-and-labels-in-ggplot2
http://www.cookbook-r.com/Manipulating_data/Converting_data_between_wide_and_long_format/
https://www.r-graph-gallery.com/48-grouped-barplot-with-ggplot2.html
https://stackoverflow.com/questions/33565199/how-to-append-multiple-files-in-r
https://stackoverflow.com/questions/44462494/include-csv-filename-when-reading-data-into-r-using-list-files
https://stackoverflow.com/questions/51957001/changing-the-order-of-the-groups-in-ggplot-in-r